---
title: "Data Preparation - Dataset Description Report"
author: "Brian Stamper"
date: "November 13, 2017"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this stage we will load all of the available data, recode variable names to more useful forms, and cast "long" data into a "wide" shape. The datasets will be organized so as to match either on SOC, CIPS, or NAICS codes. Any dataset with additional preparation applied will be built as a `data.table` and saved in native **R** `.Rdata` format.

```{r prelims, warning=FALSE}
# Package dependencies
library(readxl)
library(data.table)
if(!dir.exists('data/bls')) dir.create('data/bls')
if(!dir.exists('data/ipeds')) dir.create('data/ipeds')
if(!dir.exists('data/ohiolmi')) dir.create('data/ohiolmi')
if(!dir.exists('data/onet')) dir.create('data/onet')
if(!dir.exists('data/processed')) dir.create('data/processed')

```

### Ohio Labor Market Information

This data must be downloaded through a web interface at http://ohiolmi.com/proj/OhioJobOutlook.htm. We select area "Ohio to 2024", report "All Ohio Tables", report format "Excel". The resulting file is saved as `data/ohiolmi/Ohio2024.xlsx`.

#### Ohio Job Outlook - Occupation Detail

```{r ohiolmi-occ}

# Function to check the four-digit extension to the soc code is numeric and not all zero
is_soc6 <- function(x) {
  subsoc <- substr(x, 4, 7)
  grepl('[0-9]{4}', subsoc) & subsoc != '0000'
}

# The parameters to read_excel() were determined manually by opening the data in Excel
ohiolmi_occ <- setDT(read_excel(path = 'data/ohiolmi/Ohio2024.xlsx', 
                                sheet = 'Occupation Detail', 
                                skip = 5,
                                n_max = 756,
                                na = 'N/A'))

# Visual inspection of the data before and after the following statement is used to confirm
# the correct names are being applied to the correct columns.
names(ohiolmi_occ) <- c('soc_code_lmi',
                        'soc_desc_lmi',
                        'empl_now_lmi',
                        'occ_empl_projected_lmi',
                        'occ_empl_change_lmi',
                        'occ_empl_change_pct_lmi',
                        'ann_growth_lmi',
                        'ann_replace_lmi',
                        'ann_total_lmi',
                        'med_wage_lmi',
                        'note_lmi',
                        'ed_needed_lmi',
                        'exp_needed_lmi',
                        'otj_train_needed_lmi')

# Remove aggregation rows and keep only true SOC-6
ohiolmi_occ <- ohiolmi_occ[is_soc6(soc_code_lmi)]

# Convert annualized data to hourly - is annual if note column has '††' or '†††'
# matching note_lmi %in% c('††', '†††') doesn't work for some reason, so after checking
# one-to-one correspondence with nchar() using
# table(nchar(ohiolmi_occ$note_lmi), ohiolmi_occ$note_lmi, useNA = 'always')
# we determine we can use nchar() as a proxy to note
ohiolmi_occ[is.na(note_lmi), note_lmi := '']
ohiolmi_occ[nchar(note_lmi) %in% 2:3, med_wage_lmi := round(med_wage_lmi / 2080, 2)]

# Create factor labels
ohiolmi_occ$ed_needed_lmi <- factor(ohiolmi_occ$ed_needed_lmi,
                                        levels = c('No formal educational credential',
                                                   'Less than high school',
                                                   'High school diploma or equivalent',
                                                   'Postsecondary non-degree award',
                                                   'Some college, no degree',
                                                   'Associate\'s degree',
                                                   'Bachelor\'s degree',
                                                   'Master\'s degree',
                                                   'Doctoral or professional degree'))


ohiolmi_occ$exp_needed_lmi <- factor(ohiolmi_occ$exp_needed_lmi,
                                         levels = c('None',
                                                    'Less than 5 years',
                                                    '5 years or more'))


ohiolmi_occ$otj_train_needed_lmi <- factor(ohiolmi_occ$otj_train_needed_lmi,
                                               levels = c('None',
                                                          'Internship/residency',
                                                          'Short-term on-the-job training',
                                                          'Moderate-term on-the-job training',
                                                          'Long-term on-the-job training',
                                                          'Apprenticeship'))

# Display the data structure at this point
str(ohiolmi_occ)

# Note that there are 9/733 occupations with missing wages - will need to deal with missing later.
save(ohiolmi_occ, file = 'data/processed/ohiolmi_occ.rds')
```

#### Ohio Job Outlook - Industry Detail

```{r ohiolmi-ind}
# The parameters to read_excel() were determined manually by opening the data in Excel
ohiolmi_ind <- setDT(read_excel(path = 'data/ohiolmi/Ohio2024.xlsx', 
                                sheet = 'Industry Detail', 
                                skip = 5,
                                n_max = 251))

# Visual inspection of the data before and after the following statement is used to confirm
# the correct names are being applied to the correct columns.
names(ohiolmi_ind) <- c('naics_code_lmi',
                        'naics_desc_lmi',
                        'ind_empl_now_lmi',
                        'ind_empl_projected_lmi',
                        'ind_empl_change_lmi',
                        'ind_empl_change_pct_lmi')

# Remove aggregation rows
ohiolmi_ind <- ohiolmi_ind[!is.na(naics_code_lmi) & grepl('[0-9]+', naics_code_lmi)]

str(ohiolmi_ind)

save(ohiolmi_ind, file = 'data/processed/ohiolmi_ind.rds')
```



### O*Net Resource Center 

There is a large amount of long form data here. Casting it all wide in anticipation of the modeling phase may
in fact be inefficient and time prohibitive. Restructuring of these files will be revisited once more
detail about the needs of the modeling phase come to light.

```{r onet}
# Downloader function to save repetitive code and file transfers
onet_loader <- function(fn, ...) {
  if(!file.exists(paste0('data/onet/', fn))) {
    download.file(url = paste0('https://www.onetcenter.org/dl_files/database/db_22_1_text/', fn),
                  destfile = paste0('data/onet/', fn),
                  mode = 'wb')
  }
  fread(paste0('data/onet/', fn), na.strings = 'n/a', ...)
}

#Knowledge, Skills, Abilities
onet_knowledge <- onet_loader('Knowledge.txt')
#onet_skills <- onet_loader('Skills.txt')
#onet_abilities <- onet_loader('Abilities.txt')

# Separate the six-digit SOC and filter for cases where the suffix == 00 (the aggregate)
onet_knowledge[, soc_code := substr(`O*NET-SOC Code`, 1, 7)]
onet_knowledge[, soc_suff := substr(`O*NET-SOC Code`, 9, 10)]
onet_knowledge <- onet_knowledge[soc_suff == '00']

# rename and select relevant variables
setnames(onet_knowledge,
         old = c('Element ID',
                 'Element Name',
                 'Scale ID',
                 'Data Value'),
         new = c('elem_id',
                 'elem_name',
                 'scale_id',
                 'value'))

onet_knowledge <- subset(onet_knowledge,
                         select = c('soc_code',
                                    'elem_id',
                                    'elem_name',
                                    'scale_id',
                                    'value'))

# Cast the 'importance' next to the 'level' for each soc/knowledge
onet_knowledge <- dcast(onet_knowledge, soc_code + elem_id + elem_name ~ scale_id, value.var = 'value')




# For each occupation, we want to know if a given knowledge area is
# in the upper quartile for importance (Scale ID = 'IM') *AND* 
# competency required (Scale ID = 'LV')
# Compute the upper quantile break point:
# (using quantile because experimenting with different break points)
onet_knowledge[, high_IM := quantile(IM, .75), by = 'elem_id']
onet_knowledge[, high_LV := quantile(LV, .75), by = 'elem_id']

# Determine if both importance and level are above both for the occupation:
onet_knowledge[, high_IMLV := IM > high_IM & LV > high_LV]

# Cast the score wide over the knowledge areas:
onet_knowledge <- dcast(onet_knowledge, soc_code ~ elem_id, value.var = 'high_IMLV')

save(onet_knowledge, file = 'data/processed/onet_knowledge.rds')

onet_content_model <- onet_loader('Content%20Model%20Reference.txt')
names(onet_content_model) <- c('elem_id', 'elem_name', 'elem_desc')

save(onet_content_model, file = 'data/processed/onet_content_model.rds')


str(onet_knowledge)
str(onet_skills)
str(onet_abilities)

#Education, Experience, Training
onet_educ <- onet_loader('Education%2C%20Training%2C%20and%20Experience.txt')
onet_educ_cats <- onet_loader('Education%2C%20Training%2C%20and%20Experience%20Categories.txt')
onet_job_zones <- onet_loader('Job%20Zones.txt')
onet_job_zone_ref <- onet_loader('Job%20Zone%20Reference.txt')

str(onet_educ)
str(onet_educ_cats)
str(onet_job_zones)
str(onet_job_zone_ref)

#Interests, Work Values, Work Styles
onet_interests <- onet_loader('Interests.txt')
onet_work_values <- onet_loader('Work%20Values.txt')
onet_work_styles <- onet_loader('Work%20Styles.txt')

str(onet_interests)
str(onet_work_values)
str(onet_work_styles)

#Work Context
onet_work_context <- onet_loader('Work%20Context.txt')
onet_work_context_cats <- onet_loader('Work%20Context%20Categories.txt')

str(onet_work_context)
str(onet_work_context_cats)

```

### National Center for Education Statistics

https://nces.ed.gov/ipeds/

Integrated Postsecondary Education Data System data was downloaded through the web interface, limiting for
graduation data in Ohio. Here we only show how the data can be loaded, there is no need to re-save the data at this time.

```{r ipeds}
ipeds_complete_total <- fread('data/ipeds/CSV_1162017-305.csv', na.strings = '')
names(ipeds_complete_total) <- c('ipeds_code',
                                 'ipeds_name',
                                 'year_ipeds',
                                 'major_n_ipeds',
                                 'cip_code_ipeds',
                                 'cip_desc_ipeds',
                                 'degr_level_ipeds',
                                 'completers_n_ipeds',
                                 'IDX_C')
str(ipeds_complete_total)

ipeds_progs_offered <- fread('data/ipeds/CSV_1162017-601.csv', na.strings = '')
names(ipeds_progs_offered) <- c('ipeds_code',
                                'ipeds_name',
                                'year_ipeds',
                                'cip_code_ipeds',
                                'cip_desc_ipeds',
                                'programs_n',
                                'programs_dist_n',
                                'IDX_C')
str(ipeds_progs_offered)

ipeds_grad_rates <- fread('data/ipeds/CSV_1162017-865.csv', na.strings = '')
names(ipeds_grad_rates) <- c('ipeds_code',
                             'ipeds_name',
                             'year_ipeds',
                             'grad_rate_total_ipeds',
                             'grad_rate_pell_ipeds',
                             'grad_rate_stafford_non_pell_ipeds',
                             'grad_rate_non_stafford_non_pell_ipeds')
str(ipeds_grad_rates)

```


### Bureau of Labor Statistics 

https://www.bls.gov/jlt/

Job Openings and Labor Turnover Survey

```{r bls}
bls_loader <- function(fn, ...) {
  if(!file.exists(paste0('data/bls/', fn))) {
    download.file(url = paste0('https://download.bls.gov/pub/time.series/', fn),
                  destfile = paste0('data/bls/', fn),
                  mode = 'wb')
  }
  fread(paste0('data/bls/', fn), na.strings = 'n/a', ...)
}

bls_jolts <- bls_loader('jt/jt.data.1.AllItems')

# Extract relevant parts of the series_id
bls_jolts[, seasonal_code := substr(series_id, 3, 3)]
bls_jolts[, industry_code := substr(series_id, 4, 9)]
bls_jolts[, region_code := substr(series_id, 10, 11)]
bls_jolts[, dataelement_code := substr(series_id, 12, 13)]
bls_jolts[, ratelevel_code := substr(series_id, 14, 14)]

# fill = TRUE to handle missing varname in header, which is inconsequential
bls_jolts_dataelement <- bls_loader('jt/jt.dataelement', fill = TRUE)
bls_jolts_industry <- bls_loader('jt/jt.industry', fill = TRUE, colClasses = c(industry_code = 'character'))

# Reduce data by taking only 2016, annual, total US, by level
bls_jolts <- bls_jolts[year == '2016' & period == 'M13' & region_code == '00' & ratelevel_code == 'L']

# Get readable dataelement names
bls_jolts <- merge(bls_jolts, bls_jolts_dataelement, by = 'dataelement_code')

# Cast the data wide
bls_jolts <- dcast(bls_jolts, industry_code ~ dataelement_text, value.var = 'value')

# Merge in industry descriptors
bls_jolts <- merge(bls_jolts, bls_jolts_industry, by = 'industry_code')

# Compute a growth rate
bls_jolts[, growth := Hires / `Total separations` - 1]

str(bls_jolts)

save(bls_jolts, file = 'data/processed/bls_jolts.rds')
```

