---
title: "CPDA Machine Learning SP 2018"
author: "Brian Stamper"
date: "January 27, 2018"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 2


#### a. _Import the data into the workspace and plot it with area on x-axis and cost on y-axis._

```{r}
library(data.table)
library(ggplot2)
housing_data <- fread('data/HousingData.txt', col.names = c('area', 'cost'))
p <- ggplot(housing_data, aes(x = area, y = cost)) +
  geom_point() +
  scale_y_continuous(labels = function(x) scales::dollar(100000 * x)) +
  scale_x_continuous(labels = function(x) scales::comma(100 * x))
p
```

#### b. _Use linear least squares to find the optimal values of $\theta$._

We will set $H = \left[\begin{array}{cc}\mathbf{1} & \mathbf{area}\end{array}\right]$ and $y = \left[\begin{array}{c}\mathbf{cost} \end{array}\right]$ and compute $\theta_{LS}$ using
$$\theta_{LS} = \left(H^TH\right)^{-1}H^Ty$$

```{r}
H <- cbind(1, housing_data$area)
y <- housing_data$cost
theta_LS <- solve(t(H) %*% H) %*% t(H) %*% y
theta_LS
```

#### c. _Plot the hypothesis function $f\left(x\right)$ with the data._

```{r}
p + geom_abline(intercept = theta_LS[1], slope = theta_LS[2])
```

#### d. _Suppose if the area of the house is 700 sq. feet. Find the estimated price of the house in dollars._

To estimate the price we can use $f\left(x\right) = X^T\theta_{LS}$ where $X = \left[\begin{array}{cc}1 \\ x\end{array}\right]$

```{r}
f <- function(x) t(rbind(1, x)) %*% theta_LS
price_estimate <- scales::dollar(f(7) * 100000)   # convert to $100,000 scale
print(paste('The price estimate is', price_estimate))
```
