---
title: "CPDA Machine Learning SP 2018"
author: "Brian Stamper"
date: "January 16, 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1


```{r}
A <- matrix(c(8, -1, 2, -1, 2, -2, 2, -2, 3), 3, 3)
A
```

#### a. _Compute the determinant of the matrix A. What can you say about the rank of the matrix from the determinant?_

The determinant of A is non-zero, therefore the rows/columns are linearly independent, and A is 3x3 so the rank is min{3, 3} = 3.

```{r}
det(A)
```

#### b. _Write a code to calculate the eigenvalues and eigenvectors of the matrix A._

```{r}
ev <- eigen(A)
eigenvalues <- ev$values
eigenvectors <- lapply(1:3, function(x) ev$vectors[, x])
eigenvalues
eigenvectors
```

#### c. _Calculate the angle between the eigenvectors corresponding to the two largest eigenvalue in degrees._

Because A is symmetric, any two distinct eigenvectors should be orthogonal. (Because the first two eigenvalues are the largest, the first calculation is the actual answer, but they are all the same.)
```{r}
v1 <- eigenvectors[[1]]
v2 <- eigenvectors[[2]]
v3 <- eigenvectors[[3]]

acos(sum(v1*v2) / (sqrt(sum(v1*v1)) * sqrt(sum(v2*v2)))) * (180/pi)
acos(sum(v1*v3) / (sqrt(sum(v1*v1)) * sqrt(sum(v3*v3)))) * (180/pi)
acos(sum(v2*v3) / (sqrt(sum(v2*v2)) * sqrt(sum(v3*v3)))) * (180/pi)
```

#### d. _Is the matrix A positive definite?_

Yes, because A is symmetric and all of the eigenvalues are positive, it is positive definite.

#### e. _Let T be the matrix constructed by stacking the eigenvectors of A as its columns. Calculate  T^-1^AT._

T^-1^AT gives a diagonal matrix with the eigenvalues of A along the diagonal. (Here using "Tee" because "T" is reserved in R.)
```{r}
Tee <- matrix(c(v1, v2, v3), 3, 3)
TeeInv <- solve(Tee)
round(TeeInv %*% A %*% Tee, 6) # rounding to keep things readable
```

#### f. _Find the trace of the matrix A._

The trace of the matrix is equal to the sum of the diagonals and also is equal to the sum of the eigenvalues.
```{r}
sum(A * diag(3)) # element-wise multiplication with the identity matrix
sum(eigenvalues)
```

#### g. _Find A^2._

```{r}
A %*% A
```
